The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-27 kalman_walk_paper.tex
added overleaf edits

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}

@@

Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
gives a change of basis $P$ such that
the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
\begin{align*}
       PAP^{-1}
       &=
       \left[ \begin{array}{cccc}
           A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
           0 & A_{\ro} & 0 & A_{\ro,\nro} \\
           0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
          0 & 0 & 0 & A_{\nro}
       \end{array} \right] ,
\end{align*}
and
\begin{align*}
     PB
     &=
    \left[ \begin{array}{cccc}
         B_{\rno} \\
         B_{\ro} \\
         0 \\
         0 
   \end{array} \right] 
   &
   (CP^{-1})^T
   &=
   \left[ \begin{array}{cccc}
       0 \\
       C_{\ro}^T \\
       C_{\nrno}^T \\
       0 
   \end{array} \right] .
\end{align*}
% and
% \begin{align*}
%    CP^{-1}
%    &=
%    \left[ \begin{array}{cccc}
%        0 & C_{\ro} & C_{\nrno} & 0 
%    \end{array} \right] .
%\end{align*}
The impulse response of the system is given by
\begin{align*}
      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
\end{align*}
and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.
(Here the subscript $\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.)

@@

\paragraph{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
This leads to the question: 
How quickly does the hybrid phenotype break down with increasing distance between the parents?
To quantify this, we will measure how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
suppose that $h_0(t)$ is the \emph{optimal} impulse response function 
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
with respect to the coordinates.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}

Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Here we discuss the rate at which evolution shifts molecular mechanisms
and ask whether this process is fast enough to be a significant driver of speciation.
To do so, we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems. We work with a population of effective size $N_e$. 
\plri{note that this is more general than LTI model}

At any given time, there will be a range of regulatory coefficients present in the population
due to segregating genetic variants.
Over many generations, even if selective pressures do not change,
this range of networks will shift 
as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
%How much variation do we expect to find within a population?
%Is this range limited by available variation or kept in check by selection?
%How fast will a population explore the space of equivalent networks?
%In this section we explore informally a general model for this situation,
%in which a population drifts stochastically near a set of equivalent, optimal systems.
%We work with a population of effective size $N_e$.

Suppose that a set $x$ of coefficients that determine a system (this is $\Sys$ above),
produce a phenotype $\Phi(x)$ (the time course of $\phi(t)$).
There is an optimal phenotype $\optph$,
and a set $\optx$ of ``optimal'' coefficients that produce this phenotype.
Fitness depends on distance to the optimal phenotype --
we will write the ``distance'' between phenotypes $\phi$ and $\psi$ as $\dph(\phi, \psi)$,
measured so that the fitness of an organism with coefficents $x$ is $\fitx(x) = \exp(-\dph(\Phi(x),\optph)^2)$. 
We will assume that the map $\Phi$ is smooth
and that the optimal set $\optx$ is locally isomorphic to $\R^m$.
\plr{say that better}

%\paragraph{Offspring.}
%\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
%Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
%and the individual's coefficients are the average of her two haploid values (no dominance).
%This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
%Each new gamete is produced from the parent's two haploid copies;
%for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
%and so $\sigma$ remains constant, up to a $1/N_e$ term.
%A more general model including segregation variance \citep{barton_infinitesimal}
%would result in the same qualitative conclusions.
%\plri{but put this in the appendix?}

\paragraph{System drift}
% plr how fast does it drift
If the regulatory coefficient population variation
has standard deviation $\sigma$,
since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but mean movement along the optimal set $\optx$ is unconstrained.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depend on constraints imposed by selection and 
the covariance between genetic variation between different coefficients (the $G$ matrix \citep{G_matrix}).
% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
%For instance,
%if the variation is due to \textit{cis}-regulatory variants,
%the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
%across which a population may carry only a few common haplotypes.
%However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
%a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
%(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
%is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
%(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

% It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
% (a) a ``cloud'' of width X about the population mean, 
% which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
% In fact, the population mean will not produce exactly the optimal phenotype,
% but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.

%To obtain a general quantitative picture, we need to know 
We denote
$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
and $\gamma$, the scale on which phenotype changes moving away from $\optx$.
Concretely, $\gamma$ is the inverse of the derivative of $\dph(\Phi(x+u z), \Phi(x))$ 
with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent space at $x$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.

\paragraph{Hybridization}
The means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma^2$. 
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
These are depicted in figure \ref{fig:conceptual_fig}.

@@

Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$). As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.

@@

@@

@@

For completeness, we provide a brief exposition of how a population 
evolves due to genetic drift
with a quantitative genetics model,
as in \citet{lande1981models} or \citet{hansen1996translating}.
These ignore details of the underlying genetic basis,
but developing a more accurate model is beyond the scope of this paper.

@@

\paragraph{Completing the square}
First note that 
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right) \\
    &\qquad{}
    + \text{(terms that don't depend on $x$)} .
\end{align*}
Therefore, 
by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
\begin{align*}
    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
    &=
    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
\end{align*}

@@

Let two points  on $\optx$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
Then with $D \Phi$ and $D^2 \Phi$ the first and second derivatives of $\Phi$, respectively,
then Taylor expanding about $x_1$ and $x_2$ finds that
\begin{align*}
    \Phi(\bar x) 
    &= \Phi(x_1) + D\Phi(x_1) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) \\
    &= \Phi(x_2) - D\Phi(x_2) \cdot z + \frac{1}{2} z^T D^2 \Phi(x_2) z + O(\|z\|^3) .
\end{align*}
Now, since $\Phi(x_1) = \Phi(x_2) = \optph$ and
\begin{align*}
    D\Phi(x_2) = D\Phi(x_1) + 2 z^T D^2 \Phi(x_1) + O(\|z\|^2), \quad \text{and} \\
    D^2\Phi(x_2) = D^2\Phi(x_1) + O(\|z\|), \quad \text{and} \\
\end{align*}
adding together the two equations above and dividing by two gets that
\begin{align*}
    \Phi(\bar x) 
    &= \optph - \frac{3}{2} z^T D^2 \Phi(x_1) z + O(\|z\|^3) .
\end{align*}

\section{Differentiating the fitness function}
\label{apx:H_calc}

@@

>>>>

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}
\newcommand{\grad}{\nabla}

@@

Concretely, the \textbf{Kalman decomposition} of a system $(A,B,C)$  
gives a change of basis $P$ such that
the transformed system $(PAP^{-1},PB,CP^{-1})$  has the following form:
\begin{align*}
       PAP^{-1}
       &=
       \left[ \begin{array}{cccc}
           A_{\rno} & A_{\rno,\ro} & A_{\rno,\nrno} & A_{\rno,\nro} \\
           0 & A_{\ro} & 0 & A_{\ro,\nro} \\
           0 & 0 & A_{\nrno} & A_{\nrno,\nro} \\
          0 & 0 & 0 & A_{\nro}
       \end{array} \right] ,
\end{align*}
and
\begin{align*}
     PB
     &=
    \left[ \begin{array}{cccc}
         B_{\rno} \\
         B_{\ro} \\
         0 \\
         0 
   \end{array} \right] 
   &
   (CP^{-1})^T
   &=
   \left[ \begin{array}{cccc}
       0 \\
       C_{\ro}^T \\
       0 \\
       C_{\nro}^T
   \end{array} \right] .
\end{align*}
% and
% \begin{align*}
%    CP^{-1}
%    &=
%    \left[ \begin{array}{cccc}
%        0 & C_{\ro} & C_{\nrno} & 0 
%    \end{array} \right] .
%\end{align*}
The impulse response of the system is given by
\begin{align*}
      h(t) = C_{\ro} e^{A_{\ro} t} B_{\ro},
\end{align*}
and therefore, the system is phenotypically equivalent to the \emph{minimal} system $(A_{\ro}, B_{\ro}, C_{\ro})$.
(Here the subscript $\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.)

@@

\paragraph{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
This leads to the question: 
How quickly does the hybrid phenotype break down with increasing distance between the parents?
To quantify this, we will measure how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
suppose that $h_0(t)$ is the \emph{optimal} impulse response function 
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
with respect to the coordinates.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section*{System drift and the accumulation of incompatibilities}

Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
This begs the question: does evolution shift molecular mechanisms
fast enough to be a significant driver of speciation?
To approach this question,
we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems
due to the action of recombination, mutation, and demographic noise.
Although this is motivated by the results on linear systems above,
the quantitative genetics calculations are more general,
and only depend on the presence of genetic variation and a continuous set of phenotypically equivalent systems.

We will suppose that each organism's phenotype is determined by her vector of coefficients, denoted by $x=(x_1, x_2, \ldots, x_L)$,
and that the corresponding fitness is determined by the distance of her phenotype to optimum.
The optimum phenotype is unique, but is realized by many distinct $x$ -- those falling in the ``optimal set'' $\allS$.
The phenotypic distance to optimum of an organism with coefficients $x$ is denoted $D(x)$.
In the results above, $x = (A,B,C)$ and $D(x)$ is given by equation \eqref{eqn:distance}.
determines a system (this is a system $(A,B,C)$ above) 
with fitness determined by its phenotypic distance $D(x)$ from an optimum (this is analogous to $\allS$ above). 
Concretely, we define the fitness at $x$ to be $\exp(-D(x)^2)$.
% this assumption does not entail a loss of generality since $D$ is arbitrary.
We will assume that in the region of interest, the map $D$ is smooth
and that we can locally approximate the optimal set $\allS$ as a quadratic surface.
As above, an individual's coefficients are given by averaging her parentally inherited coefficients and adding random noise due to segregation. 
Concretely, we use the \emph{infinitesimal model} for reproduction \citep{barton_infinitesimal} --
the offspring of parents at $x$ and $x'$ will have coefficients $(x+x')/2 + \varepsilon$,
where $\varepsilon$ is a Gaussian displacement due to random assortment of parental alleles.

\paragraph{System drift}
% plr how fast does it drift
We work with a randomly mating population of effective size $N_e$. 
If the regulatory coefficient population variation
has standard deviation $\sigma$ in a particular direction,
since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but movement along the optimal set $\allS$ is unconstrained,
and so we expect the population mean to drift along the optimal set like a particle diffusing.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depends on constraints imposed by selection and 
correlations between the genetic variation underlying different coefficients (the $G$ matrix \citep{G_matrix}).
It therefore seems reasonable to coarsely model the time evolution of population variation in regulatory coefficients as 
a ``cloud'' of width $\sigma$ about the population mean, 
which moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.

% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
%For instance,
%if the variation is due to \textit{cis}-regulatory variants,
%the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
%across which a population may carry only a few common haplotypes.
%However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
%a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
%(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
%is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
%(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

There will in general be different amounts of variation in different directions;
to keep the discussion intuitive, we only discuss $\sigma_N$, the amount of variation in ``neutral'' directions
(i.e., directions along $\allS$),
and $\sigma_S$, the amount of variation in ``selected'' directions (perpendicular to $\allS$).
The other relevant scale we denote by $\gamma$,
which the scale on which distance to phenotypic optimum changes as $x$ moves away from the optimal set, $\allS$.
Concretely, $\gamma$ is
%the inverse of the derivative of 
$1/(\frac{d}{du}D(x+uz))$ 
with respect to $u$ where $x$ is optimal and $z$ is a ``selected'' direction perpendicular to $\allS$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.
Of course, there are in general many possible neutral and selected directions;
we take these values to be representative of the possible directions.

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
Each $F_2$ hybrid will be homozygous for one parental allele on average at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
but will have higher variance.
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$s will have variance of order $\sigma^2_S + \sigma^2_H T/N_e$,
where $\sigma^2_H$ is a factor that depends on the genetic basis of the coefficients.
\plr{say something about $\sigma_H$?}

What are the fitness consequences?
If a population has a Gaussian distribution of trait values with variance $\sigma^2$
and fitness decays as $\exp(-x^2/2\gamma)$ away from the population mean,
then the population mean fitness is 
\begin{align*}
   \int \frac{1}{\sigma \sqrt{2\pi}} e^{-\frac{x^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
   &=
   \sqrt{\frac{1}{1 + (\sigma/\gamma)^2}}  \le \frac{1}{\sigma/\gamma} .
\end{align*}
If $\sigma$ is much smaller than $\gamma$, the fitness consequences of variation
(i.e., genetic load) are small,
but if $\sigma$ is of order $\gamma$, then mean fitness is inversely proportional to trait variance
(i.e., $\gamma/\sigma$).
This implies that we expect $\sigma_S < \gamma$
but that once $T$ is of order $N_e (\gamma / \sigma_H)^2$, there will be loss of fitness
in $F_2$ hybrids.
Said another way,
if we write $\mathcal{F}(T)$ as the mean fitness of an $F_2$ hybrid between two populations separated by $T$ generations,
then fitness drops with the square root of $T$, measured in units of $N_e$ generations:
\begin{align*}
   \mathcal{F}(T) \le \frac{1}{\sigma_H/\gamma}\sqrt{\frac{N_e}{T}} .
\end{align*}

The fitness consequences of system drift above were due to increased \emph{variance} in $F_2$ hybrids.
If the optimal set $\allS$ is curved (which it will be in general),
then the mean $F_1$ offspring will fall away from the optimum slightly,
leading to an additional fitness cost in hybrids.
However,
as shown in appendix \ref{apx:away_from_opt} and depicted in figure \ref{fig:osc_incompat},
the deviation of the mean from the optimum grows proportionally to $T/N_e$.

\begin{align*}
\int_{-\infty}^\infty \frac{1}{\sigma\sqrt{2\pi}} e^{-\frac{(x-\mu)^2}{2\sigma^2}} e^{-\frac{x^2}{2\gamma^2}} dx
&=
\sqrt{\frac{1}{1 + \sigma^2/\gamma^2}} \exp\left\{-\frac{\mu^2}{\gamma^2}\left(\frac{1}{1 + \sigma^2/\gamma^2}\right)\right\} \\
% &\approx
% \frac{1}{\sigma/\gamma}\left(1 - \frac{\mu^2}{\gamma^2}\right) .
\end{align*}
With $\sigma = c_\sigma \gamma \sqrt{T/N_e}$
and $\mu = c_\mu \gamma (T/N_e)^2$,
this is
$$
\sqrt{\frac{1}{1 + c_\sigma^2 T/N_e}} \exp\left\{-c_\mu^2 \frac{T^2}{N_e^2} \left(\frac{1}{1 + c_\sigma^2 T/N_e}\right)\right\} \\
$$

@@

@@

%%%%%%%%%%%%
\plri{insert commented-out discussion of what the actual genetic basis is}

@@

Note that this follows directly from the fact that $F_1$ and $F_2$ hybrid phenotypes diverge from $F_0$ phenotypes, linearly and inverse-quadratically, with respect to time (more precisely $T/N_e$). As such, removing assumptions about the form of the fitness function, this model still predicts hybrid $F_2$ phenotypes to diverge much faster than $F_1$s, initially.
So even if fitness isn't as above maybe you can see this in phenotypes.

@@

For completeness, we provide a brief exposition of how a population 
evolves due to genetic drift
with a quantitative genetics model,
as in \citet{lande1981models} or \citet{hansen1996translating}.
These do not directly model underlying genetic basis,
but developing a more accurate model is beyond the scope of this paper.

@@

\paragraph{Completing the square}
First note that if $A$ is symmetric,
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so if $B$ is also symmetric and $A+B$ is invertible,
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right) \\
    &\qquad{}
    + y^T A y - y^T A^T (A+B)^{-1} A y .
\end{align*}
Since the second line doesn't depend on $x$,
by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
\begin{align*}
    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
    &=
    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
\end{align*}

@@

Let two points on $\allS$ be $x_1$ and $x_2$, let $\bar x = (x_1+x_2)/2$, and let $z=(x_2 - x_1)/2$.
Then with $\grad D$ and $\grad^2 D$ the first and second derivatives of $D$, respectively,
Taylor expanding about $x_1$ and $x_2$ finds that
\begin{align*}
    D(\bar x) 
    &= D(x_1) + \grad D(x_1) \cdot z + \frac{1}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) \\
    &=  D(x_2) - \grad D(x_2) \cdot z + \frac{1}{2} z^T \grad^2  D(x_2) z + O(\|z\|^3) .
\end{align*}
Now, since $ D(x_1) =  D(x_2) = \grad D(x_1) = \grad D(x_2) = 0$ and
\begin{align*}
    \grad D(x_2) = \grad D(x_1) + 2 z^T \grad^2  D(x_1) + O(\|z\|^2), \quad \text{and} \\
    \grad^2 D(x_2) = \grad^2 D(x_1) + O(\|z\|), 
\end{align*}
adding together the two equations above and dividing by two gets that
\begin{align*}
     D(\bar x) 
    &= \optph - \frac{3}{2} z^T \grad^2  D(x_1) z + O(\|z\|^3) .
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Gaussian load}

Suppose that a population has a Gaussian distribution in trait space
with mean $\mu$ and covariance matrix $\Sigma$,
and that fitness of an individual at $x$ is $\exp(-\|Lx\|^2/2)$.
Then, completing the square as above with $A=\Sigma^{-1}$, $y=\mu$, and $B=L^T L$,
and defining $Q = (\Sigma^{-1} + L^T L)^{-1}$,
\begin{align*}
    &
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2} x^T \Sigma^{-1} x} e^{-\frac{1}{2} x^T L^T L x} dx \\
    &\qquad =
    \frac{1}{\sqrt{2 \pi}^n \det(\Sigma)^{1/2}} 
        \int e^{-\frac{1}{2}(x-Q \Sigma^{-1} \mu)^T Q^{-1} (x-Q \Sigma^{-1} \mu)} dx 
        \\ &\qquad \qquad {}
        \times e^{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu} \\
    &\qquad =
    \sqrt{\frac{\det(Q)}{\det(\Sigma)}}
        \exp\left\{\mu^T \left( I - \Sigma^{-1} Q \right) \Sigma^{-1} \mu\right\} \\
    &\qquad =
    \sqrt{\frac{1}{\det(\Sigma)\det(\Sigma^{-1}+L^TL)}}
        \exp\left\{\mu^T \left( I - (I + L^T L \Sigma)^{-1} \right) \Sigma^{-1} \mu\right\} \\
\end{align*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Differentiating the fitness function}
\label{apx:H_calc}

@@

\plr{Add refs to sensitivity analysis.}
